%*****************************************
\chapter{Conclusions}\label{ch:conclussions2}
%*****************************************

This document showed the multiple applications that \gls{sna} have in our population within a broad range of topics. In \colorbox{PaperColor}{\textcolor{black}{Paper A}} we saw how \gls{staph} transmission is more associated with women's social interactions and how schools have distinctive strains measured by their Spa-typing. \colorbox{ResultColor}{\textcolor{black}{Result II}} also showed the importance of school interaction concerning immune response and inflammation as the academic year progresses. \colorbox{ResultColor}{\textcolor{black}{Result I}} and \colorbox{ResultColor}{\textcolor{black}{Result III}} showed the importance of friends in the quest for a healthy \gls{bmi}. \colorbox{PaperColor}{\textcolor{black}{Paper B}} shows solarium influence in women and underlying influences related to social contacts about vitamin D levels. \colorbox{ResultColor}{\textcolor{black}{Result IV}} also hints at how hormonal contraceptive usage is shared among communities and some concerning usage of \gls{otc} medicine shared in schools.

We showed the application of non-parametric simulations and machine learning methods in studying social influences. We developed an easy-to-use framework in R and showed easy-to-use examples in  \colorbox{PaperColor}{\textcolor{black}{Paper C}}, but due to several drawbacks, mainly maintenance of big libraries, is going to be ditched in favor of the Python and C++ implementations in the future.

We show several metrics to quantify these influences. However further work and discussion with health professionals are needed to determine the best way to impact policy-making based on these results at a lower cost.

In summary, this thesis provides evidence that suggests social influence is currently present in many topics, and accounting for this factor is as important, if not more, than accounting for other classical variables such as smoking or PA. We have also sadly experienced recently the necessity to account for social interactions during the SARS-Cov-2 pandemic, which also demonstrated a good example of why this type of analysis can be critical.